Short-term synaptic plasticity and threshold adaptation act in synergy to give rise to deviance detection in a spiking model of neuronal culture

### 2022-12-08 paper-1/manuscript.tex
Rev sec Methods.Paradigm

\begin{figure*}%
    \centering
    \includegraphics{fig1.pdf}
    \caption{}
    \label{fig1}
\end{figure*}

\textbf{Fig. 1}Model and paradigm. \textbf{a} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron, illustrated above the plots, firing pre-determined spikes. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Membrane voltage (black), firing threshold baseline (gray) and adaptive threshold (green) demonstrating TA. Middle: Excitatory ($g_e$, blue) and inhibitory ($g_i$, red) synaptic conductances, demonstrating excitatory STD. Bottom: Inputs to $g_e$ and $g_i$ (vertical bars in the appropriate colors), and the depression variable $x$ of the excitatory presynaptic neuron (dashed line). \textbf{b} Sample network layout, showing excitatory and inhibitory neurons and all outgoing synaptic connections from two neurons of each type. \textbf{c} Distance from stimulation site 0 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to the center of each of 5 stimulation sites (round markers), which are regularly spaced on a circle with a radius of 2.5 mm.

@@

To investigate deviance detection, we used a classical oddball paradigm as illustrated in Figure \ref{fig2}a, presenting a total of 500 stimuli at regular intervals of 500~ms. Target stimuli (labeled A hereafter) and distractor stimuli (labeled B to E) were presented in three randomized sequences: The "standard" sequence (std), consisting of 400 presentations of A and 100 presentations of B; the "deviant" sequence (dev) with 100 A and 400 B, and the many-standards control sequence (msc) with 100 presentations of each of the stimuli A through E. In this way, A is presented as the predominant and therefore expected stimulus (std), as an infrequent violation of an expectation (of B, dev), and, to control for pure adaptation effects, as an equally infrequent stimulus with no strong expectation of any particular input (msc).

@@

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig1.pdf}
    \caption{}
    \label{fig1}
\end{figure*}
\textbf{Fig. 1}Model and paradigm. \textbf{a} Membrane and synaptic dynamics of a neuron innervated by an excitatory and an inhibitory neuron, illustrated above the plots, firing pre-determined spikes. The excitatory connection has weight $w = 3$ for demonstration purposes only; a single connection with $w = 1$ is normally unable to evoke postsynaptic firing. Top: Membrane voltage (black), firing threshold baseline (gray) and adaptive threshold (green) demonstrating TA. Middle: Excitatory ($g_e$, blue) and inhibitory ($g_i$, red) synaptic conductances, demonstrating excitatory STD. Bottom: Inputs to $g_e$ and $g_i$ (vertical bars in the appropriate colors), and the depression variable $x$ of the excitatory presynaptic neuron (dashed line). \textbf{b} Sample network layout, showing excitatory and inhibitory neurons and all outgoing synaptic connections from two neurons of each type. \textbf{c} Distance from stimulation site 0 in terms of minimum number of excitatory synapses. Stimulation is provided to the 10 neurons nearest to the center of each of 5 stimulation sites (round markers), which are regularly spaced on a circle with a radius of 2.5 mm.

@@

To investigate deviance detection, we used a classical oddball paradigm as illustrated in Figure \ref{fig1}d, presenting a total of 500 stimuli at regular intervals of 500~ms. Target stimuli (labeled A hereafter) and distractor stimuli (labeled B to E) were presented in three randomized sequences: The "standard" sequence (std), consisting of 400 presentations of A and 100 presentations of B; the "deviant" sequence (dev) with 100 A and 400 B, and the many-standards control sequence (msc) with 100 presentations of each of the stimuli A through E. In this way, A is presented as the predominant and therefore expected stimulus (std), as an infrequent violation of an expectation (of B, dev), and, to control for pure adaptation effects, as an equally infrequent stimulus with no strong expectation of any particular input (msc).

@@

In order to disentangle the effects of STD and TA, we ran all simulations in four conditions: Without any short-term plasticity, with either TA only or STD only, and with both STD and TA (labeled STD+TA in the following, and corresponding to the full model as described above). To turn off STD, we replaced Equation \ref{eqn-xsyn} with a constant $x = 1$, but retained the scaling of inputs to $g_e$ with $U$ in Equation \ref{eqn-gsyn}. This corresponds to immediate recovery ($\tau_x = 0$) and was done to maintain the magnitude of the excitatory postsynaptic potential (EPSP) in the recovered state regardless of whether STD was turned on or off. To turn off TA, we replaced Equation \ref{eqn-TA} with a constant $V_{TA} = 0$, making the firing threshold completely unadaptive.
